A Stopped Negative Binomial Distribution

### 2016-07-21 stat-and-prob-letters/snb-paper.tex
Revisions and visualization for posterior underway.

%%%%%%%%%%%%%%%%%%%%%%%
%% Elsevier bibliography styles
%%%%%%%%%%%%%%%%%%%%%%%
%% To change the style, put a % in front of the second line of the current style and
%% remove the % from the second line of the style you would like to use.
%%%%%%%%%%%%%%%%%%%%%%%

%% Numbered
%\bibliographystyle{model1-num-names}

%% Numbered without titles
%\bibliographystyle{model1a-num-names}

%% Harvard
%\bibliographystyle{model2-names.bst}\biboptions{authoryear}

@@

Consider a prototypical early phase, single-arm clinical trial in which 
17 patients
are enrolled and treated. The binomial probability a patient 
responds to treatment is $p=0.2$ 
under the null hypothesis that the treatment is not effective.
If seven or more patients out of these 17 respond to the treatment then we 
reject this hypothesis and the treatment is deemed successful at 
significance level of $0.1$.  If fewer than seven respond then the null is not 
rejected and the treatment is deemed ineffective.

If all 17 patients are enrolled at once, as in the classic
design, then the sample size is 17. However, in most clinical trials the
patients are enrolled sequentially over time.
In the present example, observing seven
successful patients is one endpoint and the number of enrollees required
could be as small as seven. Similarly 11
observed treatment failures also ends the trial. This sampling mechanism, in
which the experiment ends as soon as any predefined endpoint is reached, is
called {\em curtailed sampling}. Under curtailed sampling the range of the
sample size for this trial is seven to 17.

@@

More generally the distribution of the number of trial enrollments is shown in 
Fig.~\ref{fig:exp-and-var} (a). There is relatively little probability mass
for values of seven through 10 since $p$ is small and it is unlikely 
the treatment will succeed quickly.
%The probability mass is concentrated at the 
%$11^{\text{th}}$ and $12^{\text{th}}$ step since $p$ 
%is small and for most realizations reach 11 
%non-responders before reaching 2 responders.
Fig.~\ref{fig:exp-and-var} (b) shows the expected value and variance for the
number of trial enrollments varying $p$ between zero and one. When $p$ is
small the probability of an individual response is small and the treatment
is more likely to fail shortly after the $11^{\text{th}}$ enrollment.
When $p$ is large the treatment is more likely to succeed and the number of 
enrollees approaches seven from above.

Since $p=1$ and $p=0$ represent the deterministic
case the corresponding variances are zero. The variance is
determined by the size of the support of the two endpoints and $p$, which 
weights the contribution of an endpoint based on how likely it is reached.
When $p$ is small it is more likely the trial will reach the failure endpoint
which has a support size of seven. As $p$ increases it becomes more likely the
trial reaches the success endpoint with a support size of 11.
The ``bump'' around $p=0.25$ results from inequality in the size of the
support of the two endpoints.

\begin{table}[t!]
\caption{Characteristics of the Stopped Negative Binomial Distribution}
\label{tab:snb}
\begin{center}
\begin{tabular}{|c|l|} \hline
Parameters & $p$ the success probability ($0\leq p \leq 1$; $q = 1-p$) \\
           & $s$ the number of successes before stopping ($s=1, 2, ...$)\\
           & $t$ the number of failures before stopping ($t=1, 2, ...$)\\ \hline
Support & min$(s,t) \leq k \leq s+t-1$  \\ \hline
$\mathbb{P}[Y=k]$ & ${k-1 \choose s-1} p^s (1-p)^{k-s} I_{\{s \leq k \leq s+t-1\}\}}+ {k-1 \choose t-1} (1-p)^s p^{k-t} I_{\{t \leq k \leq s+t-1\}}$\\ \hline
$\mathbb{P}[Y \leq k]$ & $2 - \mathcal{I}_{q}(k+1, s) - \mathcal{I}_{p}(k+1, t)$\\ 
    & where $\mathcal{I}$ is the regularized incomplete beta function.\\ \hline
Mean & $\frac{s}{p} \mathcal{I}_p(s,t) + \frac{p^{s-1} q^{t-1}}{B(s,t)} +
  \frac{t}{q} \mathcal{I}_q(t,s) + \frac{p^{t-1} q^{s-1}}{B(s,t)}$\\ 
  & where $B$ is the beta function \\ \hline
MGF & $\left(\frac{p e^x}{1 - qe^x}\right)^s 
  \mathcal{I}_{1-qe^x} (s, t) + \left(\frac{qe^x}{1-pe^x}\right)^t 
  \mathcal{I}_{1-pe^x}(t, s) $\\ \hline
Predictive Distribution & 
${k-1 \choose s-1} \frac{B\left(\alpha+s, k-s+\beta \right)}{B(\alpha, \beta)}+
{k-1 \choose t-1} \frac{B\left(\alpha + k-t, t+\beta\right)}{B(\alpha,\beta)}$\\
& for $p$ distributed as Beta($\alpha$, $\beta$). \\ \hline
\end{tabular}
\end{center}
\end{table}

In the rest of this work we derive the distribution of the number of 
enrollees needed
to observe either $s$ successes or $t$ failures. We refer to this distribution
as the Stopped Negative Binomial (SNB). Some of its characteristics are
summarized in Tab.~\ref{tab:snb}.
This paper derives these results
and explores properties of the distribution.
%The next section introduces our notation and basic results
%including the density of the distribution along with a description of
%its relation to other distributions. 
Section 2 derives the distribution function
based on a defined Bernoulli process and gives some basic properties.
Section 3 shows how the distribution is related to other standard
distributions and connects the SNB tail probability to the binomial tail 
probability.
Section 4 derives the moment generating function.
Section 5 derives the predictive and posterior distributions when $p$ has a 
beta distribution.

@@

To prove (\ref{eqn:pmf}), consider the
process $\mathbf{X} = \left\{X(k) : k = 0,1,... \right\}$
with $X(0)=0$ and
\begin{equation*} \label{eqn:proc}
X_{k+1} = X_k + b_{k+1} \ I_{\{ k-t < X_k < s\}}.
\end{equation*}
At each step a patient's outcome is measured. In Fig.~\ref{fig:kane_viz} 
we consider a graphical illustration of the plot $X_k$ against
$k$. If the outcome of the $k$th patient responds to treatment then the process 
advances diagonally in the positive horizontal and vertical direction. 
If the $k$th patient does not respond
then the sample path advances in the positive horizontal direction only. The
process continues until either $X_k = s$ or $X_k = k-t$.

@@

Let $\mathcal{I}_p(s, t)$ be the {\em regularized incomplete beta function} 
\citep{Olver2010}. This function satisfies 
$\mathcal{I}_p(s, t) = 1-\mathcal{I}_{1-p}(t, s)$ \citep{Uppuluri1970}. Then
\begin{align*}
R = \sum_{i=0}^{t-1} &{i+s-1 \choose i} p^s (1-p)^i +
\sum_{j=0}^{s-1}  {j+t-1 \choose j} p^j  (1-p)^t \\
   &= 1-\mathcal{I}_p(s, t) + 1 - \mathcal{I}_{1-p}(t, s) \\
   &= 1. 
\end{align*}
This completes the proof that (\ref{eqn:pmf}) is the distribution of the
stopping time and is a valid probability mass function.
\end{proof}

Next, we consider an interim analysis of a clinical trial after $s'$ 
patients respond to treatment 
and $t'$ do not for $s' < s$ and $t' < t$.
\begin{corollary} \label{conditional_distribution}
The number of subsequent enrollments needed 
to reach either endpoint behaves as SNB($p$, $s-s'$, $t-t'$).
\end{corollary}
Having observed $s'$ responders and $t'$ non-responders, there are $s-s'$ 
more responders needed to reach the success endpoint and $t-t'$ more 
non-responders needed to reach the failure endpoint.

@@

The SNB is a generalization of the negative 
binomial distribution. When $t$ is large $Y-s$ has a 
negative binomial distribution with
\begin{equation*}                                    %   (1)   
\mathbb{P}[Y=s+j\ |\ t \text{ is large\ }]        \label{nb1.eq}          
  = {{s+j-1}\choose{s-1}} p^s (1-p)^j
\end{equation*}
for $\,j=0, 1,\ldots\,$. A similar statement can be made when $s$ is large
and $t$ is small. As a result, with proper parameter choice, the SNB
can mimic other probability distributions in a manner similar to 
those described in \cite{Peizer1968} and \cite{Best1974}. Examples are
shown in in Fig.~\ref{shapes.fig}.

The SNB is the 
number of enrollment outcomes to observe {\em either} $s$ responders {\em or} 
$t$ non-responders.
The SNB generalizes both the minimum (riff-shuffle) and maximum negative
binomial distributions up to a translation of the support.
For the special case of $\,s=t,$ the distribution of $\,Y\,$ is the
riff-shuffle, or minimum negative binomial distribution~\citep{Uppuluri1970}.
The maximum negative binomial \cite{Zhang2000,Zelterman2005} is the smallest 
number of outcomes necessary to 
observe at least $s$ responders {\em and} $s$ non-responders and is equivalent
to a translated version of the riff-shuffle.

@@

%There is a close connection between the tail probabilities of the SNB and the 
%binomial distributions.
%The probability of reaching the success endpoint in an 
%SNB($p$, $s$, $t$) random variable is 
%equal to the probability of at least $s$ successes in a binomial distributed 
%random variable with size $s+t-1$ and success probability $p$. 
%Likewise, the probability of reaching the failure endpoint is equal
%to the probability of at most $s-1$ successes in a binomial distribution. 
There is an equivalence between the probability of reaching an endpoint 
in the SNB model and the tail probability of the binomial distribution.
That is, the probability that the number of responders is at least $s$ in the 
binomial model is the same as the probability the treatment succeeds 
(reaches $s$) in the SNB 
model.
\begin{prop} \label{binomial_tail}
Let $Y$ be distributed as SNB($p$, $s$, $t$) and recall that $X_Y$ corresponds
to the number of responders in a trial. Let 
$B$ be distributed binomial with size $n=s+t-1$ and response probability
$p$. Then
\begin{equation}
\mathbb{P}[B \geq s] = \mathbb{P} [X_Y = s].
\end{equation}
\end{prop}
\begin{proof}
The binomial tail probability is
\begin{equation*}
\mathbb{P}[B \geq s] = 1 - \mathcal{I}_{1-p}.
\end{equation*}
%\begin{align*}
%\mathbb{P}[B \geq s] &= \sum_{k=s}^{s+t-1} {n \choose k} p^k (1-p)^{n-k} \\
%  &= 1 - \sum_{k=0}^{s-1} {n \choose k} p^k (1-p)^{n-k} \\
%  &= 1 - \mathcal{I}_{1-p}.
%\end{align*}
The corresponding success probability is
\begin{equation*}
\mathbb{P} [X_Y = s] 
  = \sum_{k=s}^{s+t-1} {k-1 \choose s-1} p^s (1-p)^{k-s}.
\end{equation*}
Let $i=k-s$. Since
\begin{equation*}
{i+s-1 \choose s-1} = {i+s-1 \choose i},
\end{equation*}
the last summation can be rewritten as
\begin{align}
\mathbb{P} [X_Y = s] &= \sum_{i=0}^{t-1} {i+s-1 \choose i} p^s (1-p)^i\\
  &= 1 - \mathcal{I}_{1-p}(t, s)
\end{align}
completing the proof.
\end{proof}

@@

The moment generating function for the SNB is calculated in a manner similar to 
that of two negative binomial distributions. 
\begin{prop} Let $Y$ be distributed SNB with parameters $p$, $s$, and $t$.
Then the moment generating function (MGF) of $Y$ is
\begin{equation} \label{eqn:mgf}
\mathbb{E}~e^{xY} = \left(\frac{p e^x}{1 - qe^x}\right)^s 
  \mathcal{I}_{1-qe^x} (s, t) + \left(\frac{qe^x}{1-pe^x}\right)^t 
  \mathcal{I}_{1-pe^x}(t, s)
\end{equation}
for $q = 1-p$ defined for $x < \min \left\{\log(1/p), \log(1/q) \right\}$.
\end{prop}
\begin{proof}
The MGF of the SNB is:
\begin{equation*}
\mathbb{E}~e^{xY} = \sum_{k=s}^{s+t-1} {k-1 \choose s-1} p^s q^{k-s} e^{kx} 
  + \sum_{k=t}^{s+t-1} {k-1 \choose t-1} p^{k-t} q^t e^{kx}
\end{equation*}
and can be rewritten as:
\begin{equation} \label{eqn:first_sum}
\mathbb{E}~e^{xY} = \sum_{k=s}^{s+t-1}{k-1 \choose s-1} (pe^x)^{s} (qe^x)^{k-s} 
  + \sum_{k=t}^{s+t-1}{k-1 \choose t-1} (qe^x)^t (pe^x)^{k-t}.
\end{equation}
The first summation in (\ref{eqn:first_sum}) satisfies
\begin{align*}
\sum_{k=s}^{s+t-1}{k-1 \choose s-1} (pe)^{sx} (qe^x)^{k-s} &= 
  \left(\frac{pe^x}{1 - qe^x}\right)^s \ \ \sum_{k=s}^{s+t-1} {k-1 \choose s-1} 
    (qe^x)^{k-s} (1-qe^x)^s \\
  &= \left(\frac{pe^x}{1 - qe^x}\right)^s \mathcal{I}_{1-qe^x}(s, t).
\end{align*}
Since the incomplete beta function's subscript parameter has support on zero 
to one, we have $0 \leq qe^x < 1$. This also shows we must restrict
$0 \leq x < -\log(q)$.
A similar expression can be derived for the second summation in 
(\ref{eqn:first_sum}) and results in
the constraint $0 \leq x < -\log(p)$.
\end{proof}

The SNB's ability to approximate the geometric, normal, gamma, and poisson
distributions follow from it generalizing the negative binomial distribution. 
\begin{prop}
The MGF of the SNB converges to that of the negative binomial when either
$s$ or $t$ gets large. That is
\begin{equation*}
\mathbb{E} e^{xY} \rightarrow \left( \frac{pe^x}{1-qe^x} \right)^s
\end{equation*}
as $t \rightarrow \infty$. The analogous result holds when 
$s \rightarrow \infty$.
\end{prop}
\begin{proof}
The second regularized incomplete beta term in (\ref{eqn:mgf}) can be written
in terms of a cumulative binomial distribution
\begin{equation*}
\mathcal{I}_{1-pe^x}(t, s) = \mathbb{P}\left[ B \leq s-1 \right]
\end{equation*}
where $B$ is distributed as
Binomial($t-k$, $pe^x$). By Hoeffding's inequality \cite{Hoeffding1963}
it follows that
\begin{equation} \label{eqn:hoeffding}
\mathbb{P}\left[ B \leq s-1 \right] \leq 
  \exp \left(-2\frac{((t-k)pe^x - s-1)^2}{t-k} \right).
\end{equation}
As $t$ gets large $\mathcal{I}_{1-pe^x}(t, s)$ goes to zero
Likewise, $\mathcal{I}_{1-qe^x}(s, t)$ goes
to one. The proof follows by realizing 
\begin{equation*}
0 < \frac{qe^x}{1-pe^x} < 1
\end{equation*}
over the support of $x$.
\end{proof}

When $s=1$ (and $t$ is still large) the SNB's MGF is the same as that of the
geometric distribution. The negative
binomial can therefore be seen as a sum of i.i.d. geometric distributions.
For an appropriately large number of samples the central limit theorem
yields a normal approximation.

Drawing connections to the gamma and poisson distributions are more complicated.
However a connection to the gamma distribution well-studied problem in the 
literature (see \cite{Ord1968,Guenther1972,Best1974} for examples). 
A connection to the poisson appears in \cite{Anscombe1950} where
it is shown that if the mean of a poisson is proportional
to a $\chi^2_{2k}$ distribution then the negative binomial is obtained. 
Both of these approximations work by equating cumulants and then showing 
that differences between betwen the cumulant generating
functions converge to zero.

@@

Let us consider the joint SNB distribution when $p$ has a beta distribution.
\begin{prop} \label{prop:posterior}
When $p$ is distributed Beta($\alpha$, $\beta$), the 
posterior density under the SNB model is
\begin{align*}
f(p | s, t, \alpha, \beta) &= 
  \frac{ {k-1 \choose s-1} }{B(\alpha, \beta)} p^{\alpha +s -1} 
    (1-p)^{k+\beta-s-1} + \\
  & \frac{ {k-1 \choose t-1} }{B(\alpha, \beta)} p^{k+\alpha -t -1} 
    (1-p)^{\beta+t-1}.
\end{align*}
\end{prop}
\begin{proof}
This result can be found directly by multiplying the probability mass function
in (\ref{eqn:pmf}) by the density of the Beta distribution.
\end{proof}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%
%% Elsevier bibliography styles
%%%%%%%%%%%%%%%%%%%%%%%
%% To change the style, put a % in front of the second line of the current style and
%% remove the % from the second line of the style you would like to use.
%%%%%%%%%%%%%%%%%%%%%%% %% Numbered %\bibliographystyle{model1-num-names} %% Numbered without titles %\bibliographystyle{model1a-num-names} 
%% Harvard
%\bibliographystyle{model2-names.bst}\biboptions{authoryear}

@@

Consider a prototypical early phase, single-arm clinical trial in which 
17 patients
are enrolled and treated. The binomial probability of a patient 
responding to treatment is $p=0.2$ 
under the null hypothesis that the treatment is not effective.
If seven or more patients out of these 17 respond to the treatment then we 
reject this hypothesis and the treatment is deemed successful at 
significance level of $0.1$.  If fewer than seven respond then the null 
hypothesis is not 
rejected and the treatment is deemed ineffective.

If all 17 patients are enrolled at once, as in the classic
design, then the sample size is 17. However, in most clinical trials the
patients are enrolled sequentially over time.
In the present example, observing seven
successful patients is one endpoint and the number of enrollees required
could be as small as seven. Similarly 11
observed treatment failures also ends the trial. This sampling mechanism, in
which the experiment ends as soon as any predefined endpoint is reached, is
called {\em curtailed sampling}. Under curtailed sampling the range of the
sample size for this trial is seven through 17.

@@

More generally the distribution of the number of trial enrollments is shown in 
Fig.~\ref{fig:exp-and-var} (a). There is relatively little probability mass
for values of sample size equal to seven through 10 since $p$ is small and it 
is unlikely the treatment will succeed quickly.
%The probability mass is concentrated at the 
%$11^{\text{th}}$ and $12^{\text{th}}$ step since $p$ 
%is small and for most realizations reach 11 
%non-responders before reaching 2 responders.
Fig.~\ref{fig:exp-and-var} (b) shows the expected value and variance for the
number of trial enrollments varying $p$ between zero and one. When $p$ is
small then the treatment
is more likely to fail shortly after the $11^{\text{th}}$ enrollment.
When $p$ is large then the treatment is more likely to succeed and the 
number of enrollees approaches seven from above.

When $p=0$ or $1$ the processes are deterministic and variance is zero.
Value of $p$ between zero and one change the mixing proportions of 
the two endpoints. The saddle around $p=0.25$ results from inequality 
in the size of the support of the two endpoints.

%\begin{table}[t!]
%\caption{Characteristics of the Stopped Negative Binomial Distribution}
%\label{tab:snb}
%\begin{center}
%\begin{tabular}{|c|l|} \hline
%Parameters & $p$ the success probability ($0\leq p \leq 1$; $q = 1-p$) \\
%           & $s$ the number of successes before stopping ($s=1, 2, ...$)\\
%           & $t$ the number of failures before stopping ($t=1, 2, ...$)\\ \hline
%Support & min$(s,t) \leq k \leq s+t-1$  \\ \hline
%$\mathbb{P}[Y=k]$ & ${k-1 \choose s-1} p^s (1-p)^{k-s} I_{\{s \leq k \leq s+t-1\}\}}+ {k-1 \choose t-1} (1-p)^s p^{k-t} I_{\{t \leq k \leq s+t-1\}}$\\ \hline
%$\mathbb{P}[Y \leq k]$ & $2 - \mathcal{I}_{q}(k+1, s) - \mathcal{I}_{p}(k+1, t)$\\ 
%    & where $\mathcal{I}$ is the regularized incomplete beta function.\\ \hline
%Mean & $\frac{s}{p} \mathcal{I}_p(s,t) + \frac{p^{s-1} q^{t-1}}{B(s,t)} +
%  \frac{t}{q} \mathcal{I}_q(t,s) + \frac{p^{t-1} q^{s-1}}{B(s,t)}$\\ 
%  & where $B$ is the beta function \\ \hline
%MGF & $\left(\frac{p e^x}{1 - qe^x}\right)^s 
%  \mathcal{I}_{1-qe^x} (s, t) + \left(\frac{qe^x}{1-pe^x}\right)^t 
%  \mathcal{I}_{1-pe^x}(t, s) $\\ \hline
%Predictive Distribution & 
%${k-1 \choose s-1} \frac{B\left(\alpha+s, k-s+\beta \right)}{B(\alpha, \beta)}+
%{k-1 \choose t-1} \frac{B\left(\alpha + k-t, t+\beta\right)}{B(\alpha,\beta)}$\\
%& for $p$ distributed as Beta($\alpha$, $\beta$). \\ \hline
%\end{tabular}
%\end{center}
%\end{table}

In the rest of this work we derive the distribution of the number of 
enrollees needed
to observe either $s$ successes or $t$ failures. We refer to this distribution
as the Stopped Negative Binomial (SNB). 
%Some of its characteristics are summarized in Tab.~\ref{tab:snb}.
This paper derives this distribution and explores its properties.
%The next section introduces our notation and basic results
%including the density of the distribution along with a description of
%its relation to other distributions. 
Section 2 derives the distribution function
based on a defined Bernoulli process and gives some basic properties.
Section 3 shows how the distribution is related to other standard
distributions and connects the SNB tail probability to the binomial tail 
probability.
Section 4 derives the moment generating function.
Section 5 derives the predictive and posterior distributions when $p$ has a 
beta prior distribution.

@@

To prove (\ref{eqn:pmf}), consider the
process $\mathbf{X} = \left\{X(k) : k = 0,1,... \right\}$
with $X(0)=0$ and
\begin{equation*} \label{eqn:proc}
X_{k+1} = X_k + b_{k+1} \ I_{\{ k-t < X_k < s\}}.
\end{equation*}
At each step a patient's outcome is measured. In Fig.~\ref{fig:kane_viz} 
we consider a graphical illustration of the plot $X_k$ against
$k$. If the outcome of the $k$th patient responds to treatment then the process 
advances diagonally in the positive horizontal and vertical direction. 
If the $k$th patient does not respond
then the sample path advances in the positive horizontal direction only. The
process continues until either $X_k = s$ or $X_k = k-t$ corresponding to the
success and failure boundaries in Fig. \ref{fig:kane_viz}, respectively.

@@

Let $\mathcal{I}_p(s, t)$ be the {\em regularized incomplete beta function} 
\citep{Olver2010}. This function satisfies 
$\mathcal{I}_p(s, t) = 1-\mathcal{I}_{1-p}(t, s)$ \citep{Uppuluri1967}. Then
\begin{align*}
R = \sum_{i=0}^{t-1} &{i+s-1 \choose i} p^s (1-p)^i +
\sum_{j=0}^{s-1}  {j+t-1 \choose j} p^j  (1-p)^t \\
   &= 1-\mathcal{I}_p(s, t) + 1 - \mathcal{I}_{1-p}(t, s) \\
   &= 1. 
\end{align*}
This completes the proof that (\ref{eqn:pmf}) is the distribution of the
stopping time and is a valid probability mass function.
\end{proof}

Next, we consider an interim analysis of a clinical trial after $s'$ 
patients respond to treatment 
and $t'$ fail to respond for $s' < s$ and $t' < t$.
\begin{corollary} \label{conditional_distribution}
The number of subsequent enrollments needed 
to reach either $s$ or $t$ endpoints behaves as SNB($p$, $s-s'$, $t-t'$).
\end{corollary}
Having observed $s'$ responders and $t'$ non-responders, there are $s-s'$ 
more responders needed to reach the success endpoint and $t-t'$ more 
non-responders needed to reach the failure endpoint.

@@

The SNB is a generalization of the negative 
binomial distribution. When $t$ is large then $Y-s$ has a 
negative binomial distribution with
\begin{equation*}                                    %   (1)   
\mathbb{P}[Y=s+j\ |\ t \text{ is large\ }]        \label{nb1.eq}          
  = {{s+j-1}\choose{s-1}} p^s (1-p)^j
\end{equation*}
for $\,j=0, 1,\ldots\,$. A similar statement can be made when $s$ is large
and $t$ is moderate. As a result, with proper parameter choice, the SNB
can mimic other probability distributions in a manner similar to 
those described in \cite{Peizer1968} and \cite{Best1974}. Examples are
shown in in Fig.~\ref{shapes.fig}.

The SNB generalizes both the minimum (riff-shuffle) and maximum negative
binomial distributions up to a translation of the support.
For the special case of $\,s=t,$ the distribution of $\,Y\,$ is the
riff-shuffle, or minimum negative binomial distribution~\citep{Uppuluri1967}.
The maximum negative binomial \cite{Zhang2000,Zelterman2005} is the smallest 
number of outcomes necessary to 
observe at least $s$ responders {\em and} $s$ non-responders and is equivalent
to a translated version of the riff-shuffle.

@@

%There is a close connection between the tail probabilities of the SNB and the 
%binomial distributions.
%The probability of reaching the success endpoint in an 
%SNB($p$, $s$, $t$) random variable is 
%equal to the probability of at least $s$ successes in a binomial distributed 
%random variable with size $s+t-1$ and success probability $p$. 
%Likewise, the probability of reaching the failure endpoint is equal
%to the probability of at most $s-1$ successes in a binomial distribution. 
There is an equivalence between the probability of reaching an endpoint 
in the SNB model and the tail probability of the binomial distribution.
That is, the probability that the number of responders is at least $s$ in the 
binomial model is the same as the probability the treatment succeeds 
(reaches $s$) in the SNB 
model.
\begin{prop} \label{binomial_tail}
Let $Y$ be distributed as SNB($p$, $s$, $t$) and let $X_Y$ correspond
to the number of responders at the end of the trial. Let 
$B$ be distributed binomial with size $n=s+t-1$ and response probability
$p$. Then
\begin{equation}
\mathbb{P}[B \geq s] = \mathbb{P} [X_Y = s].
\end{equation}
\end{prop}
\begin{proof}
The binomial tail probability is
\begin{equation*}
\mathbb{P}[B \geq s] = 1 - \mathcal{I}_{1-p}(s, t)
\end{equation*}
%\begin{align*}
%\mathbb{P}[B \geq s] &= \sum_{k=s}^{s+t-1} {n \choose k} p^k (1-p)^{n-k} \\
%  &= 1 - \sum_{k=0}^{s-1} {n \choose k} p^k (1-p)^{n-k} \\
%  &= 1 - \mathcal{I}_{1-p}.
%\end{align*}
where $\mathcal{I}_{1-p}(s, t)$ is the {\em regularized incomplete beta} 
function. The corresponding success probability is
\begin{equation} \label{eqn:success}
\mathbb{P} [X_Y = s] 
  = \sum_{k=s}^{s+t-1} {k-1 \choose s-1} p^s (1-p)^{k-s}.
\end{equation}
Let $i=k-s$. Since
\begin{equation*}
{i+s-1 \choose s-1} = {i+s-1 \choose i},
\end{equation*}
the summation in (\ref{eqn:success}) can be rewritten as
\begin{align*}
\mathbb{P} [X_Y = s] &= \sum_{i=0}^{t-1} {i+s-1 \choose i} p^s (1-p)^i\\
  &= 1 - \mathcal{I}_{1-p}(t, s)
\end{align*}
completing the proof.
\end{proof}

@@

The moment generating function for the SNB is calculated in a manner similar to 
that of two negative binomial distributions. 
\begin{prop} Let $Y$ be distributed SNB with parameters $p$, $s$, and $t$.
Then the moment generating function (MGF) of $Y$ is
\begin{equation} \label{eqn:mgf}
\mathbb{E}~e^{xY} = \left(\frac{p e^x}{1 - qe^x}\right)^s 
  \mathcal{I}_{1-qe^x} (s, t) + \left(\frac{qe^x}{1-pe^x}\right)^t 
  \mathcal{I}_{1-pe^x}(t, s)
\end{equation}
for $q = 1-p$ and is defined for 
$x < \min \left\{\log(1/p), \log(1/q) \right\}$.
\end{prop}
\begin{proof}
The MGF of the SNB is:
\begin{equation*}
\mathbb{E}~e^{xY} = \sum_{k=s}^{s+t-1} {k-1 \choose s-1} p^s q^{k-s} e^{kx} 
  + \sum_{k=t}^{s+t-1} {k-1 \choose t-1} p^{k-t} q^t e^{kx}
\end{equation*}
and can be rewritten as:
\begin{equation} \label{eqn:first_sum}
\mathbb{E}~e^{xY} = \sum_{k=s}^{s+t-1}{k-1 \choose s-1} (pe^x)^{s} (qe^x)^{k-s} 
  + \sum_{k=t}^{s+t-1}{k-1 \choose t-1} (qe^x)^t (pe^x)^{k-t}.
\end{equation}
The first summation in (\ref{eqn:first_sum}) satisfies
\begin{align*}
\sum_{k=s}^{s+t-1}{k-1 \choose s-1} (pe)^{sx} (qe^x)^{k-s} &= 
  \left(\frac{pe^x}{1 - qe^x}\right)^s \ \ \sum_{k=s}^{s+t-1} {k-1 \choose s-1} 
    (qe^x)^{k-s} (1-qe^x)^s \\
  &= \left(\frac{pe^x}{1 - qe^x}\right)^s \mathcal{I}_{1-qe^x}(s, t).
\end{align*}
Since the regularized incomplete beta function's subscript parameter 
has support on zero 
to one, we have $0 \leq qe^x < 1$. This implies
$x < -\log(q)$.
A similar expression can be derived for the second summation in 
(\ref{eqn:first_sum}) and results in
the constraint $x < -\log(p)$.
\end{proof}

The SNB's ability to approximate the geometric, normal, gamma, and Poisson
distributions follow from it generalizing the negative binomial distribution. 
\begin{prop}
The MGF of the SNB converges to that of the negative binomial when either
$s$ or $t$ gets large. That is
\begin{equation*}
\mathbb{E} e^{xY} \rightarrow \left( \frac{pe^x}{1-qe^x} \right)^s
\end{equation*}
as $t \rightarrow \infty$. The analogous result holds when 
$s \rightarrow \infty$.
\end{prop}
\begin{proof}
The second regularized incomplete beta term in (\ref{eqn:mgf}) can be written
in terms of a cumulative binomial distribution
\begin{equation*}
\mathcal{I}_{1-pe^x}(t, s) = \mathbb{P}\left[ B \leq s-1 \right]
\end{equation*}
where $B$ is distributed as
Binomial($t-k$, $pe^x$). By Hoeffding's inequality \cite{Hoeffding1963}
it follows that
\begin{equation} \label{eqn:hoeffding}
\mathbb{P}\left[ B \leq s-1 \right] \leq 
  \exp \left(-2\frac{((t-k)pe^x - s-1)^2}{t-k} \right).
\end{equation}
As $t$ gets large $\mathcal{I}_{1-pe^x}(t, s)$ goes to zero
Likewise, $\mathcal{I}_{1-qe^x}(s, t)$ goes
to one. The proof follows by realizing 
\begin{equation*}
0 < \frac{qe^x}{1-pe^x} < 1
\end{equation*}
over the support of $x$.
\end{proof}

When $s=1$ (and $t$ is still large) the SNB's MGF is approximately 
the same as that of the
geometric distribution. The negative
binomial can therefore be seen as a sum of i.i.d. geometric distributions.
For an appropriately large number of samples, the central limit theorem
yields a normal approximation.

Drawing connections to the gamma and Poisson distributions are more complicated.
However a connection to the gamma distribution well-studied problem in the 
literature (see \cite{Ord1968,Guenther1972,Best1974} for examples). 
A connection to the Poisson appears in \cite{Anscombe1950} where
it is shown that if the mean of a Poisson is proportional
to a chi-square distribution with $2k$ degrees of freedom
then the negative binomial is obtained. 
Both of these approximations work by equating cumulants and then showing 
that differences between betwen the cumulant generating
functions converge to zero.

@@

Let us consider the posterior distribution of $p$ with an SNB likelihood 
and Beta($\alpha$, $\beta$) prior distribution.
The prior times the likelihood is given by the function
\begin{align} \label{eqn:ptl}
f_p(x | \alpha, \beta ) f_{Y|p}(x | k, s, t) &= 
  \frac{ {k-1 \choose s-1} }{B(\alpha, \beta)} x^{\alpha +s -1} 
    (1-x)^{k+\beta-s-1} + \\
  & \frac{ {k-1 \choose t-1} }{B(\alpha, \beta)} p^{k+\alpha -t -1} 
    (1-x)^{\beta+t-1} \nonumber
\end{align}
where $0 \leq x \leq 1$, $s \leq k \leq s+k-1$, and $t \leq k \leq s+k-1$.
This result can be found directly by multiplying the probability mass function
in (\ref{eqn:pmf}) by the density of the Beta distribution.

The predictive distribution of the SNB can be found by integrating
$x$ over the interval zero to one and applying the definition of the 
beta function.
\begin{align} \label{eqn:predictive}
\int_0^1 f_p(x | \alpha, \beta )  f_{Y|p}(x | k, s, t) & dx = 
  {k-1 \choose s-1} \frac{B\left(\alpha+s, k-s+\beta \right)}{B(\alpha, \beta)} 
    + \nonumber \\
  & {k-1 \choose t-1} 
    \frac{B\left(\alpha + k - t, t+\beta\right)}{B(\alpha, \beta)}.
\end{align}

The ratio of (\ref{eqn:ptl}) divided by (\ref{eqn:predictive}) gives the 
posterior distribution of $p$. It is a mixture of beta distributions. The
mixing parameter depend on the endpoints ($s$ and $t$), the number of enrollees needed to reach an endpoint ($k$), and the prior parameters ($\alpha$ and
$\beta$).

%\begin{proof}
%For notational simplicity, assume that $k$ is in the range
%$\min(s,t) \leq k \leq s+t-1$. When 
%this is not the case appropriate terms should be removed as dictated by the indicator functions.
%\begin{align*}
%f(k | s, t, \alpha, \beta) = \frac{1}{B(\alpha, \beta)} & \int_0^1 {k-1 \choose s-1} p^{\alpha +s -1} \left(1-p\right)^{k-s+\beta-1} + \\
% & {k-1 \choose t-1} p^{k-t+\alpha-1}\left(1-p\right)^{t+\beta-1} dp \\
%= \frac{1}{B(\alpha, \beta)}  {k-1 \choose s-1} & \int_0^1  p^{\alpha +s -1} \left(1-p\right)^{k-s+\beta-1} dp + \\
% & \frac{1}{B(\alpha, \beta)} {k-1 \choose t-1} \int_0^1  p^{k-t+\alpha-1}\left(1-p\right)^{t+\beta-1} dp.
%\end{align*}
%The result in (\ref{eqn:posterior}) follows by the definition of the 
%beta function.
%\end{proof}
